% !TEX root = ../main.tex

\vspace{20pt}
I would first like to thank my partner Sophie for her support and understanding through this
process and for all the sacrifices she has made on my behalf.  I would similarly like to
thank my family for their unwavering support and making me who I am today.  I would like
to thank Frank Wood, Jan-Willem van de Meent, and Brooks Paige for their guidance and
support, both academic and personal.  I owe them all, particularly Frank and Jan-Willem,
a huge debt of gratitude for teaching me most of what I know, nurturing me from an
arrogant fool who asked questions like \emph{Metropolis what?} to an even more arrogant
fool who thinks he knows everything.  This work would have been a terrible failure without them.
I would like to thank my friends and colleagues 
Atılım Güneş Baydin, Rob Cornish, Piotr Czaban, Neil Dhir, Jack Fitzsimons, Adam Goliński,
Bradley Gram-Hansen, Max Igl, David Janz, Tom Jin, Tuan Anh Le, Mario Lezcano Casado, 
Aravindh Mahendran, David Martínez Rubio, Siddharth N, Nantas Nardelli, Michael Osborne,
Nick Palmius, Yura Perov, David Tolpin, Andrea Vedaldi, Andrew Warrington, Stefan Webb, 
Hongseok Yang, Yuan Zhou, and Rob Zinkov, for making this time in Oxford some of the 
happiest of my life.  I would like to thank my previously unmentioned coauthors Arnaud Doucet, Fredrik Lindsten,
Christian A. Naesseth, and Benjamin Vincent, for being an absolute pleasure to work with.
I would like to thank Bradley Gram-Hansen, Tuan Anh Le, Mario Lezcano Casado, Sophie Madden, Siddharth N, Rif Saurous,
Andrew Warrington, Frank Wood, and Yuan Zhou for providing feedback on earlier drafts of this thesis.
Finally, I would like to thank BP for providing the funding for this research.

Much of the novel material in this thesis is from coauthored work and it would be fabrication
to claim it all as my own.  In particular, a lot of the work is based around the probabilistic 
programming system \emph{Anglican}, for which all the credit must go to Frank Wood,
Jan-Willem van de Meent, and David Tolpin. Elsewhere, citations are made
to the original coauthored papers upon which this work is based, for which credit
must go to all my coauthors.