\documentclass[a4paper,twoside,12pt]{article}

\include{preamble}
\usepackage{etoolbox}
\usepackage{hyperref}
% \patchcmd{<cmd>}{<search>}{<replace>}{<success>}{<failure>}
\patchcmd{\abstract}{Abstract}{Overview}{}{}

\thispagestyle{plain}

\ShortHeadings{Declaration of Contribution}{Tom Rainforth}
\firstpageno{1}

\setlength{\parindent}{0em}
\setlength{\parskip}{5pt}

\begin{document} 

\title{\vspace{-40pt} Declaration of Contribution for D.Phil Thesis}

\author{\name Tom Rainforth \email twgr@robots.ox.ac.uk\\
	        \addr 
	        Department of Engineering Science \\
	         University of Oxford}

%\editor{}
\maketitle


\vspace{-30pt}

Much of the work presented in this thesis originates from joint authored publications.  The declaration
outlines my contributions to those papers, with all deviations in the thesis from the published work being my own.

 
\section{Interacting Particle Markov Chain Monte Carlo}

\textit{\bibentry{rainforth2016interacting}}

I was co-lead author along with Christian.  The initial idea was developed concurrently by a number of the
coauthors but not including myself.  Development of the idea into the final algorithm was spearheaded by
myself, Christian, and Fredrik.  I carried out all of the code implementation and experiments.  Writing was carried
out roughly evenly between myself and Christian will help from the other authors.  The main proof was done by
Christian with help from Fredrik, Arnaud, and Brooks.

\section{Bayesian Optimization for Probabilistic Programs}

\textit{\bibentry{rainforth2015workshopbopp}}

\textit{\bibentry{rainforth2016nips}}

I was the lead author and carried developed the vast majority of the technical innovations myself, though Jan-Willem, Frank,
and Mike all also had input in the original idea, while Jan-Willem and Frank assisted in the
development the key marginal transformation.  The code implementation was lead by myself with assistance from
Jan-Willem and Tuan Anh, the latter of which did the low level implementations of the code transformations based on my
designs.  Experiments were a joint effort between myself, Jan-Willem, and Tuan Anh.  I did the majority of the writing,
but had assistance from Jan-Willem, Tuan Anh, and Frank.  

\section{Nested Monte Carlo}

\textit{\bibentry{rainforth2016pitfalls}}

\textit{\bibentry{rainforth2017pitfalls}}

I was the lead author and developed the initial idea.  I provided the high level ideas for all of the theorems along with
informal proofs, with were then developed into more formal proofs by myself, Rob, and Hongseok.  I took the
lead on formalizing (using the numbering of~\cite{rainforth2017pitfalls}) Theorems 2, 5, and 7 myself, Rob took the lead on Theorems 1, 3 and 4, and
Hongseok took the lead on Theorem 6.  Experiments were carried out myself, except for the cancer simulations which were assisted by Andrew.
The derivation of the Bayesian experimental design was done by myself.  Writing was lead by myself with assistance from
the other authors, in particular Rob.

\section{Bayesian Experimental Design}

\textit{\bibentry{vincent2017darc}}

All machine learning and statistical contributions (derivation of estimator, design of inference scheme, development of
optimizer etc) were done by myself, with Ben developing the model and applications specifics.  The code implementation
was a joint venture, with myself carrying out the technical components of the code and providing the core code structure,
but with Ben doing the experiments and adapting the code base further.

\section{Anglican Probabilistic Programming System}

My contribution to the Anglican PPS itself (other that with BOPP) is relatively limited.  Credit for Anglican goes primarily
to David Tolpin, Jan-Willem van de Meent, and Frank Wood.

\section{Other Papers}

The thesis cites other papers for which I am coauthor, but does not include any substantial material from
the original papers (some sentences have been duplicated, but only those which were written by myself in the original work).
%
%\textit{\bibentry{rainforth2015canonical}}
%
%\textit{\bibentry{janz2016probstruct}}
%
%\textit{\bibentry{le2017auto}}
%
%\textit{\bibentry{warrington2017Opt}}

\nobibliography{refs} 

\end{document} 

