% !TEX root = ../main.tex

\chapter{Probabilistic Programming -- the User's Perspective}
\label{chp:probprog}

Probabilistic programming systems (PPSs) allow users to define probabilistic models 
using a domain-specific programming language or modeling library. A probabilistic program implicitly 
defines a distribution on random variables, whilst the system back-end implements 
general-purpose inference methods.  Two key challenges for a PPS are in providing the
syntax and semantics to allow easy definition of models, and in designing the solvers, i.e.
inference engines, to provide effective inference for those models.
In this chapter, we focus on the former of these, providing an introduction to 
probabilistic programming from a user's perspective, explaining how it can be used for, and
to extend, conventional Bayesian modeling and how it can also be used to reinterpret many computational simulation
techniques not usually associated with Bayesian modeling in a more Bayesian mindset.
We will mostly ignore the rather major issue of how to construct automated inference engines for probabilistic
programs, returning to address this in Chapter~\ref{chp:proginf}.
Here we will explain how general purpose PPSs aim to  provide 
the \emph{flexibility} to define
wide ranging and potentially obscure models, the \emph{expressivity} of a framework for 
model definition that is more in line we conventional scientific simulation than mainstream 
statistical approaches, and the \emph{automation} to  run any problem the user might write
by decoupling model specification and inference.
Together these characteristics produce a framework that allows researchers whose expertise 
lies elsewhere, to carry out powerful statistical analyses for their application specific tasks.  
This framework also aids in the development of both inference
algorithms and models for those within the machine learning and statistics communities,
by removing many of the complications of one while developing the other.
%
%We will for now focus on a user's perspective, introducing how and why one might want to use
%probabilistic programming, explain how ideas from more mainstream Bayesian
%modeling can be transferred, and demonstrate how probabilistic programming can be
%used to both extend the traditional Bayesian framework and reinterpret many computational simulation
%techniques not usually associated with Bayesian modeling in a more Bayesian mindset.
%We will mostly ignore the rather major issue of how to conduct Bayesian inference for
%models defined in probabilistic programs, assuming for now that we have a magical
%general purpose inference engine that will solve any model we provide,
%before returning to actually confront this major stumbling
%block in Chapter~\ref{chp:proginf} once we have introduced the problems of Bayesian
%inference more generally.  We will provide a brief introduction to the Anglican PPS
%to give us a platform for providing examples and highlighting
%key components of designing a PPS.  Finally, we discuss some of the current limitations
%and opportunities for PPSs (other than the obvious computational
%issues which we return to later), in particular demonstrating how we can use them to go beyond
%standard inference frameworks and why we would want to do so.

We note that it will be necessary at times during this chapter to refer briefly to some Bayesian inference
algorithms that will not be properly introduced until Chapters\ref{chp:inf} and~\ref{chp:part}.  
We have situated this chapter before those partly in order to emphasize the point that one should not 
need an intricate knowledge of inference methods to \emph{use} PPSs.  Though it is difficult
to introduce PPSs while completely omitting reference to inference methods, readers who
are not familiar with inference methods should be able to safely ignore which methods are referred to
at a first pass, noting only that different inference algorithms have different requirements and sets 
of problems they perform well on, and thus that the design of a PPS is intricately linked to the inference
method(s) used.

\input{probprog/sim.tex}
\input{probprog/two.tex}
\input{probprog/models.tex}

\section{An Introduction to Anglican}
\label{sec:probprog:anglican}

One such general purpose system, \emph{Anglican}, will be used as a reference in this paper.  In Anglican, models are defined using the inference macro \defquery. These models, which we refer to as queries \citep{goodman_uai_2008}, specify a joint distribution $p(Y,X)$ over data $Y$ and variables $X$. Inference on the model is performed using the macro \doquery, which produces a sequence of approximate samples from the conditional distribution $p(X|Y)$ and, for importance sampling based inference algorithms (e.g. sequential Monte Carlo), a marginal likelihood estimate $p(Y)$.  

Random variables in an Anglican program are specified using \sample statements, which can be thought of as terms in the prior. Conditioning is specified using \observe statements which can be thought of as likelihood terms.  Outputs of the program, taking the form of posterior samples, are indicated by the return values.  There is a finite set of \sample and \observe statements in a program source code, but the number of times each statement is called can vary between executions.  We refer the reader to  \href{http://www.robots.ox.ac.uk/~fwood/anglican/}{\small\url{http://www.robots.ox.ac.uk/~fwood/anglican/}} for more details.
%
%\section{Going Beyond Bayesian Inference}
%\label{sec:probprog:limit}

\todo[inline]{
Interpreters rather than custom languages

Avoiding bugs in code}